---
title: "DBSCAN: Introduction & general usage guide"
output: rmarkdown::html_vignette
bibliography: 'dbscan-intro-references.bib'
vignette: >
  %\VignetteIndexEntry{DBSCAN: Introduction & general usage guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.height = 5,
  fig.width = 7,
  comment = "#>"
)
```

# Introduction

DBSCAN - short for "Density-based Clustering Application with Noise" - is a density-based non-parametric clustering algorithm first proposed by @dbscan-orig. The general idea is that clusters are formed when observations are grouped together in a "dense" manner; this "density" can be measured by how many of them are located within a certain range/distance of each other. There are two inter-related criteria implied by this measurement, and they are the most critical parameters of the algorithm:

* $min\_pts$: minimum number of points required to be found within $\epsilon$ distance

* $\epsilon$: the maximum distance in which to find $min\_pts$ number of points
    
Points that fulfil both of this criteria are what is called "core points", as they will form the *core* of the generated clusters. Clusters are then expanded to include other points - called "border points" - which may not fulfil the $min\_pts$ criteria by itself, but is nevertheless within $\epsilon$ distance of a core point. Any remaining points which does not fulfil both criteria will be considered as "noise" - i.e., does not belong to any cluster.

The abstract algorithm of DBSCAN is as follows [@dbscan-revisit]:

1. Compute neighbors of each point, and identify core points.

2. Join neighbouring core points into initial clusters.

3. For non-core points, determine their reachability from a core point.

4. Assign to a cluster if reachable; otherwise assign as noise.
    
The two main advantages of DBSCAN are [@dbscan-when]:

1. There is no need to pre-specify the number of clusters to create: core points are automatically found based on the given $min\_pts$ and $\epsilon$, and the number of core point groups will correspond to the eventual created clusters.

2. Ability to find irregularly-shaped clusters: assigning non-core points to a cluster is similar to finding another point that already belong to a cluster, regardless of where they are within ϵ. This is unlike, for example, k-means clustering where points are assigned based on their direct distance to the centroids, leading to clusters that are generally circle-like.

# Usage

```{r setup}
library(dbscanDATA501)
```

For this introduction, two 2-D toy datasets will be used:

* `blobs` have observations grouped into two distinct approximately-globular clusters.

    ```{r}
    blobs <- read.csv('dat/blobs.csv')
    head(blobs)
    
    plot(blobs, pch=19, xlim=c(-15, 10), ylim=c(-10, 15))
    ```

* `circles` have observations grouped into two hollow circles inside each other.

    ```{r}
    circles <- read.csv('dat/circles.csv')
    head(circles)
    
    plot(circles, pch=19, xlim=c(-1.1, 1.1), ylim=c(-1.1, 1.1))
    ```

Now let's apply the DBSCAN algorithm to the `blobs` dataset using the `dbscan()` function. The initial parameters will be set to $min\_pts = 5$ and $\epsilon = 0.2$. This means a cluster will be initiated when at least one point has at least 5 other points within 0.2 units of it's range.

```{r}
blobs_dbscan <- dbscan(blobs, 0.2, 5)
```

The `dbscan()` functions returns a few things, but the most important is `cluster_labs`: an integer vector indicating to which cluster each observation belongs to, with 0 indicating noise. Note that the exact numbering of the cluster labels is arbitrary; the important thing is points labelled 1 belong to one cluster, points labelled 2 belong to another cluster, etc.

Now the original data can be plotted with the addition of the newly-formed clusters to display which points belong to which cluster:

```{r}
plot(blobs, pch=19, xlim=c(-15, 10), ylim=c(-10, 15),
     col=factor(blobs_dbscan$cluster_labs))
```

Two distinct clusters have been formed, with noise points in black.

Next, let's do the same thing with `circles` with $\epsilon = 0.1$:

```{r}
circles_dbscan <- dbscan(circles, 0.1, 5)
plot(circles, pch=19, xlim=c(-1.1, 1.1), ylim=c(-1.1, 1.1),
     col=factor(circles_dbscan$cluster_labs))
```

Two clusters have been generated, with each hollow circle forming it's own cluster

# References
